Monitoring and Projecting Global Hunger

### 2020-12-10 docs/modeling_fies.tex

\author[1,2,*]{Matthew Cooper}
\author[2,3]{Benjamin Müller}
\author[4]{Carlo Cafiero}
\author[2,5]{Juan Carlos Laso Bayas}
\author[5,6]{Jesús Crespo Cuaresma}
\author[2,7]{Homi Kharas}

\affil[1]{T.H. Chan School of Public Health, Harvard University}
\affil[2]{The World Data Lab, Vienna, Austria}
\affil[3]{Department of Economics, Vienna University of Economics and Business}
\affil[4]{The Food and Agricultural Organization of the United Nations}
\affil[5]{The International Institute for Applied Systems Analysis}
\affil[6]{Wittgenstein Centre For Demography and Global Human Capital}
\affil[7]{The Brookings Institute}
\affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}

\maketitle
\begin{abstract}
Using microdata from 75 countries collected by the FAO and Gallup World Poll for the Voices of the Hungry project, we modeled levels of food security at the subnational level from 2010 to 2030 to create the \href{https://worldhunger.io}{World Hunger Clock}.  This is the first global picture at a subnational level of the Food Insecurity Experience Scale, the food security metric most indicative of the lived experience of hunger.  We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population food insecure in some high-income countries to the parts of developing world where over half the population is severely food insecure.  Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of food insecurity have been steadily declining, with large observed and forecasted gains in lower and lower-middle income countries, and decreases in total number of severely food insecure people.  However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, will continue to increase through the end of the 2020s.  Overall, we conclude that global gains have been incremental, and current trends in development and demographic change will still leave a large share of the world's population experiencing hunger by 2030.  \end{abstract}

@@

Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the UN \cite{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \citep{wambogo2018validity, smith2017world}.  The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.  Using a Rasch model \citep{Cafiero2018}, each individual in the survey is given a severity score that maximizes the likelihood of observing the responses given to the 8 questions asked in the survey.  Once individuals are ranked in this fashion, thresholds for moderate and severe food insecurity are set that describe people's experiences, and the prevalence of such food insecurity events can be computed. A moderately food insecure adult is likely to have compromised on food quality and variety as a result of lack of income, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally. A severely food insecure adult is likely to have run out of food and has gone an entire day without eating at times during the year.  Because the model estimates the population over a threshold on a univariate food insecurity scale, the percentage of people over the threshold for moderate food insecurity includes people that are also over the threshold for severe food insecurity.

@@

\section{Results}
\subsection{Food Insecurity over Time}
Globally, in the year 2020, we estimate that 820 million people are above the threshold for severe food insecurity, and 2.5 billion people are above the threshold for moderate food insecurity (See Fig. \ref{fig:timeseries}). These numbers are higher than the estimates in the 2020 FAO State of Food Insecurity report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \citep{sofi2020}.

@@

Examining the rate of food insecurity in a population, rather than the total number of food insecure people (See Fig. \ref{fig:rates}, we find clear improvements in all except high income countries.  Countries that are currently upper-middle income, like China and Brazil, saw large gains in improving food insecurity rates throughout the 2010s, and will continue to see somewhat diminished gains in improving food security throughout the 2030s.  Lower and lower-middle income countries, on the other hand, saw more modest gains in the 2010s but are forecasted to see greater gains in reducing the rate of food insecurity in 2020s.  In spite of this progress in the bottom three income quartiles, in high-income countries, the rates of people in moderate and severe food insecurity have been flat and are expected even to increase slightly.  Overall, the world has made steady but slow progress on reducing the rate of food insecurity.

@@

For severe food insecurity, the most important variable for predicting food security levels is the Poverty Headcount Index (See Fig. \ref{fig:vimp}, followed by the rate of stunting.  For predicting the rate of people with at least moderate food insecurity, GDP per capita is the most important, followed closely by the Poverty Headcount Index.  For the rate of severe food insecurity, the country's Gini coefficient was more important compared to moderate food insecurity.  Across both indicators, economic variables were the most relevant predictors.

@@

The results of this analysis can be further explored in more detail on the \href{https://worldhunger.io}{World Hunger Clock}, including statistics for each subnational administrative area.

\section{Discussion}
This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts until the year 2030.  We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the coronavirus pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in developing countries.  Nevertheless, lower and lower-middle income countries are expected to see overall increases in the number of food insecure people.

@@

When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and hunger are the same across all locations.  We test the accuracy of these cases by using ten-fold cross validation at the country level, and we find good accuracy, with our models making estimates that are only off by 4-7 percentage points on average when applied to countries outside of the training set.

@@

To get the best possible model performance, it is necessary to tune the hyperparamters, which control the process in which the random forest algorithm is run.  These include the average number of observations in a leaf node, the number of variables randomly selected as candidates for splitting a node, and the maximum depth of each tree.  We use the combination of hyperparameters that perform best under ten-fold cross validation based on $R^2$.

@@

After fitting the models, we examine the importance of the individual covariates in explaining rates of moderate and severe food insecurity. We use a common method of permuting each individual variable and re-running the regressions, and then determining the difference in Mean Average Error (MAE) between the regression \citep{ishwaran2007variable, breiman2001random}.  This metric can provide some interpretability to model predictions.

>>>>

\author[1,2,*]{Matthew Cooper}
\author[2,3]{Benjamin Müller}
\author[4]{Carlo Cafiero}
\author[2,5]{Juan Carlos Laso Bayas}
\author[3,5,6,7]{Jesús Crespo Cuaresma}
\author[2,8]{Homi Kharas}

\affil[1]{T.H. Chan School of Public Health, Harvard University}
\affil[2]{The World Data Lab, Vienna, Austria}
\affil[3]{Department of Economics, Vienna University of Economics and Business}
\affil[4]{Statistics Division, Food and Agricultural Organization of the United Nations}
\affil[5]{International Institute for Applied Systems Analysis}
\affil[6]{Wittgenstein Centre for Demography and Global Human Capital}
\affil[7]{Austrian Institute of Economic Research}
\affil[8]{The Brookings Institute}
\affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}

\maketitle
\begin{abstract}
Using microdata from 75 countries collected by FAO through the Gallup World Poll for the Voices of the Hungry project, we modeled levels of food security at the subnational level from 2010 to 2030 to create the World Hunger Clock (https://worldhunger.io).  This is the first global picture of food security at a subnational level, based on the Food Insecurity Experience Scale, a food security metric developed by FAO that is most indicative of the individual or household lived experience of food insecurity and hunger.  We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population being at least moderately food insecure in some high-income countries to parts of the developing world where over half the population is severely food insecure.  Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of food insecurity have been declining steadily yet slowly, with larger observed and forecasted gains in lower and lower-middle income countries, and global decreases in the total number of severely food insecure people.  However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, will continue to increase through the end of the 2020s.  Overall, we conclude that global gains have been incremental, and current trends in development and demographic change will still leave a large share of the world's population experiencing hunger by 2030.  \end{abstract}

@@

Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the United Nations (UN) \cite{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \citep{wambogo2018validity, smith2017world}.  The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.  Using a Rasch model \citep{Cafiero2018}, each individual in the survey is given a severity score that maximizes the likelihood of observing the responses given to the 8 questions asked in the survey.  Once individuals are ranked in this fashion, thresholds for moderate and severe food insecurity are set that describe people's experiences, and the prevalence of such food insecurity events can be computed. A moderately food insecure adult is likely to have compromised on food quality and variety as a result of lack of resources, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally. A severely food insecure adult is likely to have run out of food and has gone an entire day without eating at times during the year.  Because the model estimates the population over a threshold on a univariate food insecurity scale, the percentage of people over the threshold for moderate food insecurity includes people that are also over the threshold for severe food insecurity.

@@

\section{Results}
\subsection{Food Insecurity over Time}
Globally, in the year 2020, we estimate that 820 million people are above the threshold for severe food insecurity, and 2.5 billion people are above the threshold for moderate food insecurity (See Fig. \ref{fig:timeseries}). These numbers are higher than the estimates in the 2020 FAO State of Food Insecurity (SOFI) report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \citep{sofi2020}.

@@

Examining the rate of food insecurity in a population, rather than the total number of food insecure people (See Fig. \ref{fig:rates}), we find clear improvements in all except high income countries.  Countries that are currently upper-middle income, like China and Brazil, saw large gains in improving food insecurity rates throughout the 2010s, and will continue to see somewhat diminished gains in improving food security throughout the 2030s.  Lower and lower-middle income countries, on the other hand, saw more modest gains in the 2010s but are forecasted to see greater gains in reducing the rate of food insecurity in 2020s.  In spite of this progress in the bottom three income quartiles, in high-income countries, the rates of people in moderate and severe food insecurity have been flat and are expected even to increase slightly.  Overall, the world has made steady but slow progress on reducing the rate of food insecurity.

@@

For severe food insecurity, the most important variable for predicting food security levels is the Poverty Headcount Index (See Fig. \ref{fig:vimp}), followed by the rate of stunting and the Gini coefficient.  For predicting the rate of people with at least moderate food insecurity, GDP per capita is the most important, followed closely by the Poverty Headcount Index.  For the rate of severe food insecurity, the country's Gini coefficient was more important compared to moderate food insecurity.  Across both indicators, economic variables were the most relevant predictors.

@@

The results of this analysis can be further explored in more detail on the World Hunger Clock \href{https://worldhunger.io}{(worldhunger.io)}, including statistics for each subnational administrative area.

\section{Discussion}
This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts until the year 2030.  We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the COVID-19 pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in developing countries.  Nevertheless, lower and lower-middle income countries are expected to see overall increases in the number of food insecure people.

@@

When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and hunger are the same across all locations.  We test the accuracy of these cases by using leave-one-out cross validation at the country level, and we find good accuracy, with our models making estimates that are only off by 4-7 percentage points on average when applied to countries outside of the training set.

@@

To get the best possible model performance, it is necessary to tune the hyperparamters, which control the process in which the random forest algorithm is run.  These include the average number of observations in a leaf node, the number of variables randomly selected as candidates for splitting a node, and the maximum depth of each tree.  We use the combination of hyperparameters that perform best under leave-one-out validation based on $R^2$.

@@

After fitting the models, we examine the importance of the individual covariates in explaining rates of moderate and severe food insecurity. We use a common method of permuting each individual variable and re-running the regressions, and then determining the difference in MAE between the regression \citep{ishwaran2007variable, breiman2001random}.  This metric can provide some interpretability to model predictions.
